---
title: "Project Time Series"
author: "Carolin Drenda, Silje Anfindsen, Jonathan Stålberg"
date: "3 5 2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

````{r}
serie=ts(read.table("CO2IndUSA.dat"),start=1990,freq=12)
plot(serie, type = "o")
abline(v=1990:2020,col=4,lty=3)
```

# Identification

## a)
```{r}
m=apply(matrix(serie,nr=12),2,mean)
v=apply(matrix(serie,nr=12),2,var)
plot(m,v,xlab="Mean",ylab="variance",main="serie")
#abline(lm(v~m), col=2, lty=3,lwd=2)
```


```{r}
boxplot(serie~floor(time(serie)))
```





```{r}
monthplot(serie)
```



```{r}
d12serie <- diff(serie, 12)

```


```{r}
plot(d12serie, type = "o")
abline(h=0)
```

```{r}
d1d12serie <- diff(d12serie)

```


```{r}
plot(d1d12serie, type = "o")
abline(h=0)

```

```{r}
var(serie)
var(d12serie)
var(d1d12serie)
var(diff(d1d12serie)) # not needed
```
## b) 
```{r}
par(mfrow = c(1,2))
acf(d1d12serie, ylim = c(-1,1), lag.max = 60, col = c(2,rep(1,11)), lwd = 2)
pacf(d1d12serie, ylim = c(-1,1), lag.max = 60, col = c(rep(1,11),2), lwd = 2)

```


Seasonal: MA(1),(AR(3))
Regular: ARMA(1,1), Ar(2), MA(1) (2,3)

# Estimation


## a)
Model 1
```{r}
(mod=arima(d1d12serie, order=c(1,0,1),seasonal=list(order=c(0,0,1),period=12)))

```
mean non-sign

```{r}
# (mod1=arima(serie, order=c(1,1,1),seasonal=list(order=c(0,1,1),period=12)))
# ar 1 not sign
(mod1=arima(serie, order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12)))

```


```{r}
validation=function(model,dades){
  s=frequency(get(model$series))
  resid=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resid,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resid),3*sd(resid)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resid)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resid)
  qqline(resid,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resid,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resid),sd=sd(resid)),col=2,add=T)
  
  ## Individual Correlation Tests 
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resid,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #ACF & PACF of square residuals 
  par(mfrow=c(1,2))
  acf(resid^2,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid^2,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #Global Correlation Test
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=72)
  names(psis)=paste("psi",1:72)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:20])
  
  plot(psis,type="h",main="Pesos Psis - MA infinito")
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=72)
  names(pis)=paste("pi",1:72)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:20])

  plot(pis,type="h",main="Pesos Pis - AR infinito")
  
#   #Some Complementary Tests
#   cat("\nNormality tests\n")
#   cat("\n--------------------\n")  
#   ##Shapiro-Wilks Normality test
#   print(shapiro.test(resid(model)))
# 
#   suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
# ##Anderson-Darling test: Normality
#   print(ad.test(resid(model)))
#   
#   suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
#   ##Jarque-Bera test: Normality
#   print(jarque.bera.test(resid(model)))
#   
#   cat("\nHomoscedasticity Test\n")
#   cat("\n--------------------\n")
#   suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
#   ##Breusch-Pagan test
#   obs=get(model$series)
#   print(bptest(resid(model)~I(obs-resid(model))))
#   
#   cat("\nIndependence Tests\n")
#   cat("\n--------------------\n")
#   
#   ##Durbin-Watson test
#   print(dwtest(resid(model)~I(1:length(resid(model)))))
#   
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resid(model),type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  
  
  #Sample ACF vs. Teoric ACF: similar?
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  acf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample ACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36),ylim=c(-1,1), 
       type="h",xlab="Lag",  ylab="", main="ACF Teoric")
  abline(h=0)
  
  #Sample PACF vs. Teoric PACF
  pacf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample PACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36, pacf=T),ylim=c(-1,1),
       type="h", xlab="Lag", ylab="", main="PACF Teoric")
  abline(h=0)
  par(mfrow=c(1,1))
}

```

```{r}
validation(mod1, d1d12serie)
```
Ljung box is sign -> look for different model

```{r}
(mod2 = arima(d1d12serie, order=c(2,0,0),seasonal=list(order=c(0,0,1),period=12)))
```
intercept not
```{r}
#BEST MODEL:
(mod3 = arima(serie, order=c(2,1,0),seasonal=list(order=c(0,1,1),period=12)))
```
# Validation

## a,b,d)
```{r}
validation(mod3,d1d12serie)
```

# c)Stability
```{r}
ultim = c(2018,12)
#pdq = c(2,1,0)
#PDQ = c(0,1,1)
serie1=window(serie)  
serie2=window(serie,end=ultim)         
```

```{r}
(mod31= arima(serie1, order=c(2,1,0),seasonal=list(order=c(0,1,1),period=12)))
(mod32= arima(serie2, order=c(2,1,0),seasonal=list(order=c(0,1,1),period=12)))
```
Problem: shorter TS has lower aic


# Prediction

# a)
```{r}
pred=predict(mod32,n.ahead=12)
pr<-ts(c(tail(serie2,1),pred$pred),start=ultim,freq=12)

se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(pr-1.96*se,start=ultim,freq=12)
tu<-ts(pr+1.96*se,start=ultim,freq=12)
pr<-ts(pr,start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-3,+2),type="o",main="Model ARIMA(1,1,1)(0,1,1)12")
abline(v=(ultim[1]-3):(ultim[1]+2),lty=3,col=4)
```
```{r}
obs=window(serie,start=ultim)
(mod.EQM1=sqrt(sum(((obs-pr)/obs)^2)/12))
(mod.EAM1=sum(abs(obs-pr)/obs)/12)
```
```{r}
pred <- predict(mod31,n.ahead=12)
pr<-ts(c(tail(serie1,1),pred$pred),start=ultim + c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim + c(1,0),freq=12)

#Intervals
tl1<-ts(pr-1.96*se,start=ultim + c(1,0),freq=12)
tu1<-ts(pr+1.96*se,start=ultim + c(1,0),freq=12)
pr1<-ts(pr,start=ultim + c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main="Model ARIMA(2,1,0)(0,1,1)12")
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
previs1=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0))
```


# Outlier treatment

```{r}
source("CalendarEffects.r")
source("atipics2.r")
```

## a) Calendar effect:
```{r}

data=c(start(serie)[1],start(serie)[2], length(serie)) #starting year, month, series size
(wTradDays=Wtrad(data)) #creates auxiliary variable for trading days configurations (5/2 the ideal proportion)
```
```{r}
(wEast=Weaster(data))
```
```{r}
(mod3EC=arima(serie,order=c(2,1,0),seasonal=list(order=c(1,1,0),period=12),xreg=data.frame(wTradDays,wEast)))
```
```{r}
(mod3Ea=arima(serie,order=c(2,1,0),seasonal=list(order=c(1,1,0),period=12),xreg=data.frame(wEast)))
```
```{r}
(mod3TD=arima(serie,order=c(2,1,0),seasonal=list(order=c(1,1,0),period=12),xreg=data.frame(wTradDays)))
```
Only use trading days not easter, because lowest aic and other parameters are not significant

Calculate Trading day effect:
```{r}
EfecTD=coef(mod3TD)["wTradDays"]*wTradDays
plot(EfecTD)
```

corrected series for trading day effect and compare with original serie:
```{r}
serieTD=serie-EfecTD
plot(serie)
lines(serieTD,col=2)
```


Transform into stationarity:
```{r}
d12serieTD <- diff(serieTD, 12)
d1d12serieTD <- diff(d12serieTD)
```

Variance
```{r}
var(serieTD)
var(d12serieTD)
var(d1d12serieTD)
var(diff(d1d12serieTD)) # not necessary
```
Same Transformation as before.


```{r}
par(mfrow=c(1,2), mar =c(1,2,4,1))
acf(d1d12serieTD,ylim=c(-1,1),lag.max=72,col=c(2,rep(1,11)),lwd=2)
pacf(d1d12serieTD,ylim=c(-1,1),lag.max=72,col=c(rep(1,11),2),lwd=2)
```
Modelidentification:
seasonal: MA(1) (AR(3))
Regular: MA(2), AR(2)


Mdoel estimation
```{r}
#(modec <- arima(serie, order=c(0,1,1),seasonal=list(order=c(0,1,2),period=12), xreg=data.frame(wTradDays) ))
# sma 2 not sign
(modec <- arima(serie, order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12), xreg=data.frame(wTradDays)))
```
```{r}
# same parameters as before
(modec2 <- arima(serie, order=c(2,1,0),seasonal=list(order=c(1,1,0),period=12), xreg=data.frame(wTradDays) ))
```
```{r}
(modec3 <- arima(serie, order=c(2,1,0),seasonal=list(order=c(0,1,1),period=12), xreg=data.frame(wTradDays) ))
```

Prefer modec ARIMA(0,1,1)(0,1,1)_12 + wTradDays since aic is smaller,
modec3 with ARIMA(2,1,0)(0,1,1)_12 + wTradDays also possible



```{r}
dades=d1d12serieTD
model=modec
validation(model,dades)
```



```{r}
dades=d1d12serieTD
model=modec3
validation(model,dades)
```

Prefer modec3 (arima(2,1,0)(0,1,1)) because there are only 2 sign Ljung Box

Maybe change model and use modec instead of modec3





Effect of trading days:
```{r}
EfecTD=coef(modec)["wTradDays"]*wTradDays
plot(EfecTD)
```

## b) Outlier
(session 5:)
```{r}

# automatic detection of outliers
mod.atip=outdetec(modec,dif=c(1,12),crit=2.8,LS=T)

# estimated variance after outlier detection and treatment
mod.atip$sigma2

modec$sigma2

```
Variance after detection is smaller

```{r}
#Table with detected outliers, their types, magnitud, statistic values and chronology
atipics=mod.atip$atip[order(mod.atip$atip[,1]),]

meses=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")

data.frame(atipics,Fecha=paste(meses[(atipics[,1]-1)%%12+1],start(serie)[1]+((atipics[,1]-1)%/%12)))
```
```{r}
#additional column: percentage variation
data.frame(atipics,Fecha=paste(meses[(atipics[,1]-1)%%12+1],start(serie)[1]+((atipics[,1]-1)%/%12)),perc.Obs=exp(atipics[,3])*100)
```

Linearize serie, (without outliers)
```{r}
serie.lin=lineal(serie,mod.atip$atip)
#serieEC.lin=serie.lin-EfecTD
```

```{r}
plot(serie.lin,col=2, ylim = c(100,160))
lines(serie)
```
Plot of outlier profile
```{r}
plot(serie-serie.lin)
```

```{r}

d12serie.lin=diff(serie.lin,12)
d1d12serie.lin=diff(d12serie.lin)
plot(d1d12serie,col=1,type="o")
abline(v=1990:2020,lty=3,col=4)
abline(h=0)
```
Identify model:

```{r}
par(mfrow=c(1,2))
acf(d1d12serie.lin,ylim=c(-1,1),lag.max=72,col=c(2,rep(1,11)),lwd=2)
pacf(d1d12serie.lin,ylim=c(-1,1),lag.max=72,col=c(rep(1,11),2),lwd=2)
```
Seasonal: MA(1) MA(2) (AR(3))
Regular: AR(1), AR(2), MA(1)


```{r}

#(mod.lin2=arima(serie.lin,order=c(2,1,0),seasonal=list(order=c(0,1,2),period=12),xreg=data.frame(wTradDays)))
#ma 2 not sign
(mod.lin2=arima(serie.lin,order=c(0,1,1),seasonal=list(order=c(0,1,1),period=12),xreg=data.frame(wTradDays)))
```
Looks okay:

```{r}
dades=d1d12serie.lin
model=mod.lin2
validation(model,dades)
```




```{r}
(mod.lin=arima(serie.lin,order=c(2,1,0),seasonal=list(order=c(3,1,0),period=12),xreg=data.frame(wTradDays)))
dades=d1d12serie.lin
model=mod.lin
validation(model,dades)
```
This model does not look goof

Stability
```{r}
ultim=c(2018,12)
pdq=c(0,1,1)
PDQ=c(0,1,1)

serie1=window(serie.lin,end=ultim+c(1,0))
serie2=window(serie.lin,end=ultim)

wTradDays2=window(wTradDays,end=ultim)


(mod=arima(serie1,order=pdq,seasonal=list(order=PDQ,period=12),xreg=data.frame(wTradDays)))
(mod2=arima(serie2,order=pdq,seasonal=list(order=PDQ,period=12),xreg=data.frame(wTradDays2)))

```
model is stable


## Forecast:


Insample prediction: (not in assignment?)
```{r}
pred=predict(mod.lin2,n.ahead=12,newxreg=window(wTradDays,start=c(ultim[1]+1,1)))
predic=pred$pr

wLS=sum(mod.atip$atip[mod.atip$atip$type_detected=="LS" & mod.atip$atip$Obs<=length(serie)-12,3])
pr<-ts(c(tail(serie2,1),predic)+wLS,start=ultim,freq=12)


se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(pr-1.96*se,start=ultim,freq=12)
tu<-ts(pr+1.96*se,start=ultim,freq=12)
pr<-ts(pr,start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-3,2),type="o",main="Model ARIMA(2,1,0)(3,1,0)12+CE+Atip")
abline(v=(ultim[1]-3):(ultim[1]+2),lty=3,col=4)


```
red obs in 2019 is weird, besides that: looks good, 

```{r}
obs=window(serie,start=ultim)
(mod.EQM3=sqrt(sum(((obs-pr)/obs)^2)/12))

(mod.EAM3=sum(abs(obs-pr)/obs)/12)
```



```{r}
data3=c(ultim[1]+2, 1, 12)

wTradDays3=Wtrad(data3)

pred=predict(mod.lin2,n.ahead=12,newxreg=data.frame(wTradDays3)) ##wEast3))
predic=pred$pr
wLS=sum(mod.atip$atip[mod.atip$atip$type_detected=="LS",3])
pr<-ts(c(serie[length(serie)],predic+wLS),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

#Intervals
tl3<-ts(pr-1.96*se,start=ultim+c(1,0),freq=12)
tu3<-ts(pr+1.96*se,start=ultim+c(1,0),freq=12)
pr3<-ts(pr,start=ultim+c(1,0),freq=12)

ts.plot(serie,tl3,tu3,pr3,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,3),type="o",main="Model ARIMA(0,1,1)(0,1,1)12+CE+Atip")
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
previs3=window(cbind(tl3,pr3,tu3),start=ultim+c(1,0))
```
## Compare models


```{r}
par(mfrow = c(1,2))
ts.plot(serie,previs1,lty=c(1,2,1,2),col=c(1,4,2,4),xlim=c(2018,2021),type="o",main="Model ARIMA(2,1,1)(0,1,1)12")
abline(v=2016:2021,lty=3,col=4,ylim=c(15,280))
ts.plot(serie,previs3,lty=c(1,2,1,2),col=c(1,4,2,4),xlim=c(2018,2021),type="o",main="Model ARIMA(3,1,0)(2,1,0)12+EC+Atip")
abline(v=2016:2021,lty=3,col=4,ylim=c(15,280))
```

```{r}
resul=data.frame(
  par=c(length(coef(mod3)),length(coef(mod.lin2))+nrow(mod.atip$atip)),
  
  Sigma2Z=c(mod3$sigma2, mod.lin2$sigma2),
  AIC=c(AIC(mod3), AIC(mod.lin2) + 2 * nrow(mod.atip$atip)),
  BIC=c(BIC(mod3), BIC(mod.lin2) + log(length(serie)-13) * nrow(mod.atip$atip)),
  RMSPE=c(mod.EQM1,mod.EQM3),
  MAPE=c(mod.EAM1,mod.EAM3),
  meanLength=c(sum(previs1[,3]-previs1[,1]),sum(previs3[,3]-previs3[,1]))/12)
row.names(resul)=c("ARIMA(2,1,0)(0,1,1)12","ARIMA(0,1,1)(0,1,1)12+EC+IA+Atip")


resul
```
